A thesis draft

### 2018-11-24 res/sections/08-Background.tex

\section{Virtualization}
It is remarkable how cloud computing has surged in the recent years. This is due
the numerous features it carries, such as a great way to isolate processes and
mitigate possible attacks from malicious users. On top of that, it allows to run
``virtual'' machines with different configurations, without bringing any
changes to the host machine. Indeed, the virtualization concepts refer to the
property of a machine to be virtualized. To achieve this, the components that
forms a common computer need to be virtualized too. Thus, these assets are
recreated in this virtualized manner, e.g. Storage devices and computer network
resources~\cite{liu2014research}. At the time writing, there are two common
virtualization options: the virtualization through Hypervisor (defined also as
hypervisor-based) and the container virtualization.

\paragraph*{Hypervisor}
The first one is known for providing a very good isolation from the host
machine~\cite{eder2016hypervisor}, even if studies found possibilities for an
attacker to escape from the a virtual machine and operate on the host
machine\footnote{An example can be found in
  \url{https://github.com/MorteNoir1/virtualbox_e1000_0day}}. Even though every
component in a hypervisor environment is virtualized, there are possibilities for
the ``guest'' OS (the virtualized one), to access to physical devices (like USB
keys attached on the ``host'' OS).

\paragraph*{Container-based}
Container-based virtualization uses a different approach to reach isolation.
Instead of creating a whole new stack, it applies very strict policies to the
programs running on the ``host'' machine. This concept has its roots back to the
jail environments, where group of process where isolated by the main OS, that
provided a virtualized file-system~\cite{canonico2007virtualization}. In this
context, a container-based applications can be seen as a jailed application with
policies aimed to protect the host machine.

@@

Virtualization assumes another context when referred to networking: virtualized
networking allow multiple, heterogeneous networks to cohabit, increasing the
flexibility and manageability. This network separation creates two different
entities, that becomes decouples: who provides the infrastructure (commonly
called InPs) and who provides the services (SPs). A virtual network can be
create aggregating one or more InPs, thus creating a virtual network end-to-end
service~\cite{chowdhury2009network}. An example of this are the Virtual Private
Networks (VPNs), that are able to cross public networks, giving access to
content that otherwise would not be available locally.

\section{Software-defined networking}
The concept of Software-defined networking (SDN) is not new: starting from
1996~\cite{sezer2013we} it aims to provide user-controlled management of
forwarding in network nodes. This means to have the possibility to control the
forwarding procedures. A close examples of SDN implementations are Ethane and
OpenFlow, respectively created in 2007 and 2008. In an SDN architecture, there
is the decoupling of data and control plane\todo{ref}, but there is a
centralization of the networking logic and state~\cite{fundation2012software}.
The use of an SDN network is estimated to reduce the Capital Expenditure (CapEX)
and Operational Expenditure (OpEX) costs~\cite{benzekki2016software}.

@@

As the Figure~\ref{chap:background:img:sdn_architecture} illustrates, the SDN
architecture is formed by three layers, defined as planes, that accomplish
specific tasks~\cite{fundation2012software} as
in~\cite{fundation2013software}. Below we will perform a brief explanation of
each plane.

@@

\subsubsection{Control plane}
The SDN controllers, as just said in the previously, receives requests from the
SDN applications. Since the controller is centralized, it can therefore exploit
the complete knowledge of the network. This gives the possibility to apply
networking optimization like flow management~\cite{sezer2013we}. The controller
abstracts from the network complexity, and it can collect information about the
network performing requests to the \emph{southbound} interfaces. The control
plane can have multiple controllers, that are in turn federated thanks to the
\emph{eastbound} and \emph{westbound} (e.g.
HyperFlow)~\cite{benzekki2016software}, that allow several existing controller
written with different languages to coexists.

\subsubsection{Data Plane}
The data plane is the lowest plane, and it provides the aforementioned
\emph{southbound} API. This kind of API can be deployed in two different kind of
scenarios: for in-band communications and for out-of-band
communications~\cite{benzekki2016software}. The data plane gives access to the
physical and virtual switches, routers and access-points, and it gives the
possibility to perform programmatic forwarding operations, but it is also able
to provide statistic reporting and it supports event notifications.

\subsection{OpenFlow protocol}
OpenFlow is an open communication protocols that enables the possibility to
actually program networks, offering the possibility to change the flow-tables
for layer 3 switches and for routers~\cite{mckeown2008openflow}. In other words,
the OpenFlow protocol gives the possibility for controller to access at these
components. Routes can be changed on the typology of the packet, or on the state
of the network (i.e. if a network is under heavy load, certain packages can be
discarded).

@@

\section{Network slicing}
Internet Service Providers (ISP) provides networking services to a wide range of
users, that have different necessities between them, and that can require
different kind of characteristics, e.g. a network connectivity in a train has
different networking requirements from the one regarding a company. It is with
these presuppositions that emerges the necessity of service optimization based
on the necessities of the users. To solve this problem in the recent years the
concept of Networks Slice (NS) has been introduced~\cite{alliance20155g}. The
NS aims to logically separate this set of services that present different
use-cases, and to achieve this it exploits the SDN and Network Function Chaining
(NFC) that allows network softwarization~\cite{ordonez2017network}.

@@

Even if the first idea of NS goes back to 2002~\cite{peterson2003blueprint}, NSs
ancestor can be found in VLANs, that allows different hosts with no physical
connectivity to be in the same network. The NS concept has attracted a lot of
interest especially from the Industries, that is trying to introduce them in
the next mobile network standard, the 5G. More details are provided in
Section~\ref{chap:background:sec:5g:sub:ns}.

@@

\paragraph*{Network Functions}
The concept behind NF is to usually elaborate incoming data and forward it to
destination. NF can be physical of virtualized: in the first case is commonly a
set of hardware, in the second case instead it can be a software implementation,
decoupled from the hardware where its running. The Virtualized Network Functions
(VNFs), that are recently emerging propose a multitude of challenges, especially
regarding their latency, reliability and efficiency due their virtualized nature
of network function.

@@

\subsection{Network Function Virtualization}
Concept recently developed in 2012 by the ETSI organization, the Network
Function Virtualization (NFV), is a network architecture concept that aims, with
the virtualization technologies available in the market, to virtualize network
nodes. It is believed that NFV will transform the way the global communications
will work, thanks to its disruptive trend in the networking
industry~\cite{gray2016network}. Essentially, a NFV, working in harmony with the
NS aims to provide sufficient packet processing efficiency, especially in
environment such as the cloud one.

@@

The MANagement and Orchestration (MANO) architecture is divided in turn in three
main components, such as the Virtual Infrastructure Manager (VIM), the VNF
Manager (VNFM) and the NFV Orchestrator (NFVO). The MANO organize and holds
important data, regarding the Network Service (NS) catalog, the list of the VNFs
definition (VNF catalog) and the current NFV instances and the NFVI resources.

@@

The SPH contains, more than the 24-bit Service Path Identifier (SPI), a Service
Index (SI), that provides the location within the SFP.

\subparagraph*{Context header}
There are two kind of context header definition: one with a fixed context header
length (16-Byte) and one with a dynamic context header length.

\noindent Regarding this field, the standard does not make any assumption about
the content, and it is left free to be filled with custom metadata. The only
requirement is to set this filed with zero when is not used.

@@

\section{The upcoming connectivity standard: 5G}
The continuous innovation in the mobile network connectivity is leading to the
creation of a new standard, the 5G, which is estimated to arrive to the market
consumer in 2020~\cite{iwamura2015ngmn}. Lead by the Next Generation Mobile
Network (NGMN) alliance, a group composed by the major players in the field of
mobile connectivity, the 5G aims to offer not only at the end-user a new way to
browse the web, download and watch interactive content but also to create an
ad-hoc solution for Machine-to-Machine (M2M) data traffic, which is increasing
more than ever thanks to the spreading of IoT devices, which sensors need to
continuously send data to servers/data-centers. Relatively to \emph{Long Term
  Evolution} (LTE), 5G points to have data rates $10$ times better, with $10$
times smaller end-to-end latency and an increased connection density by $100$
times~\cite{alliance20155g}.

>>>>

\section{Virtualization}
It is remarkable how cloud computing has surged in the recent years. This is due
the numerous features it carries, such as a great way to isolate processes and
mitigate possible attacks from malicious users. On top of that, it allows to run
virtual machines with different configurations, without bringing any changes to
the host machine. Indeed, the virtualizations concepts refer to the property of
a machine to be virtualized. To achieve this, the components that forms a common
computer need to be virtualized too. Thus, these assets are recreated in this
virtualized manner, e.g. storage devices and computer network
resources~\cite{liu2014research}. At the time writing, there are two common
virtualization options: the virtualization through Hypervisor (defined also as
hypervisor-based) and the container virtualization.

\paragraph*{Hypervisor-based}
The first one is known for providing a very good isolation from the host
machine~\cite{eder2016hypervisor}, even if studies found possibilities for an
attacker to escape from the a virtual machine and operate on the host
one\footnote{An example can be found in
  \url{https://github.com/MorteNoir1/virtualbox_e1000_0day}}. Even though every
component in a hypervisor environment is virtualized, there are possibilities
for the ``guest'' OS (the virtualized one), to access to physical devices (like
USB keys attached on the ``host'' OS).

\paragraph*{Container-based}
Container-based virtualization uses a different approach to reach isolation.
Instead of creating a whole new stack, it applies very strict policies to the
programs running on the ``host'' machine. This concept has its roots back to the
jail environments, where group of process where isolated by the main OS, that
provided a virtualized file-system without any possibility to access to
different paths~\cite{canonico2007virtualization}. In this context, a
container-based application can be seen as a jailed application with policies
aimed to protect the host machine.

@@

Virtualization assumes another context when referred to networking: virtualized
networking allows multiple, heterogeneous networks to cohabit, increasing the
flexibility and manageability. This network separation creates two different
entities, that becomes decoupled: who provides the infrastructure (commonly
called InPs) and who provides the services (SPs). A virtual network can be
created aggregating one or more InPs, thus creating a virtual network end-to-end
service~\cite{chowdhury2009network}. An example of this are the Virtual Private
Networks (VPNs), that are able to cross public networks, giving access to
content that otherwise would not be available locally.

\section{Software-defined networking}
The concept of Software-defined networking (SDN) is not new: starting from
1996~\cite{sezer2013we} it aims to provide user-controlled management of data
forwarding in network nodes. This means to have the possibility to control the
forwarding procedures. A close examples of SDN implementations are Ethane and
OpenFlow, respectively created in 2007 and 2008. In an SDN architecture, there
is the decoupling of data and control plane (which will be explained in
Section~\ref{chap:background:sec:sdn:sub:cp}), but there is a centralization of
the networking logic and state~\cite{fundation2012software}. The use of an SDN
network is estimated to reduce the Capital Expenditure (CapEX) and Operational
Expenditure (OpEX) costs~\cite{benzekki2016software}.

@@

As the Figure~\ref{chap:background:img:sdn_architecture} illustrates, the SDN
architecture is formed by three layers, defined as planes, that accomplish
specific tasks~\cite{fundation2012software} as explained
in~\cite{fundation2013software}. Below we will perform a brief explanation of
each plane.

@@

\subsubsection{Control plane}
\label{chap:background:sec:sdn:sub:cp}
The SDN controllers, as just said in the previously, receives requests from the
SDN applications. Since the controller is centralized, it can therefore exploit
the complete knowledge of the network. This gives the possibility to apply
networking optimization like flow management~\cite{sezer2013we}. The controller
abstracts from the network complexity, and it can collect information about the
network performing requests to the \emph{southbound} interfaces. The control
plane can have multiple controllers, that are in turn federated thanks to the
\emph{eastbound} and \emph{westbound} (e.g.
HyperFlow)~\cite{benzekki2016software} interfaces, that allow several existing
controller written even with different languages to coexists.

\subsubsection{Data Plane}
The data plane is the lowest plane and it provides the aforementioned
\emph{southbound} API. This API can be deployed in two different kind of
scenarios: for in-band communications and for out-of-band
communications~\cite{benzekki2016software}. The data plane gives access to the
physical and virtual switches, routers and access-points, and it gives the
possibility to perform programmatic forwarding operations, but it is also able
to provide statistic reporting and it supports event notifications.

\subsection{OpenFlow protocol}
OpenFlow is an open communication protocols that enables the possibility to
actually program networks, offering the possibility to change the flow-tables
for layer 3 switches and for routers~\cite{mckeown2008openflow}. In other words,
the OpenFlow protocol gives the possibility for controller to access at these
components. Routes can be changed based on the typology of the packet, or based
on the state of the network (i.e. if a network is under heavy load, certain
packages can be discarded).

@@

\section{Network slicing}
Internet Service Providers (ISP) provides networking services to a wide range of
users, that have different necessities between them, and that can require
different kind of characteristics, e.g. a network connectivity in a travelling
train has different networking requirements from the one deployed in a company.
It is with these presuppositions that emerges the necessity of service
optimization based on the users necessities. To solve this problem in the recent
years the concept of Networks Slice (NS) has been
introduced~\cite{alliance20155g}. The NS aims to logically separate this set of
services that present different use-cases, and in order to achieve this it
exploits the SDN and Network Function Chaining (NFC) that allows network
softwarization~\cite{ordonez2017network}.

@@

Even if the first idea of NS goes back to 2002~\cite{peterson2003blueprint}, NSs
ancestor can be found in VLANs, that allows different hosts with no physical
connectivity to be in the same network. The NS concept has attracted a lot of
interest especially from the Industries, that is trying to introduce them in the
5G mobile network standard. About this, more details are provided in
Section~\ref{chap:background:sec:5g:sub:ns}.

@@

\paragraph*{Network Functions}
The concept behind NF is to usually elaborate incoming data and forward it to
destination. NFs can be physical of virtualized: in the first case is commonly a
set of hardware devices, in the second case instead it can be a software
implementation, decoupled from the hardware where its running. The Virtualized
Network Functions (VNFs), that are recently emerging propose a multitude of
challenges, especially regarding their latency, reliability and efficiency due
their virtualized nature of network function.

@@

\subsection{Network Function Virtualization}
Concept recently developed in 2012 by the ETSI organization, the Network
Function Virtualization (NFV), is a network architecture concept that aims, with
the virtualization technologies available in the market, to virtualize network
nodes. It is believed that NFV will transform the way the global communications
will work, thanks to its disruptive trend in the networking
industry~\cite{gray2016network}. Essentially, a NFV, working in harmony with the
NS aims to provide sufficient packet processing efficiency, especially in
environments such as the cloud one.

@@

The MANagement and Orchestration (MANO) architecture is divided in turn in three
main components, such as the Virtual Infrastructure Manager (VIM), the VNF
Manager (VNFM) and the NFV Orchestrator (NFVO). The MANO organize and holds
important data, regarding the Network Service (NS) catalog, the list of the VNFs
definitions (VNF catalog) and the current NFV instances and the NFVI resources.

@@

The SPH contains the 24-bit Service Path Identifier (SPI) and a Service Index
(SI), that provides the location within the SFP.

\subparagraph*{Context header}
There are two kind of context header definitions: one with a fixed context
header length (16-Byte) and one with a dynamic context header length.

\noindent Regarding this field, the standard does not make any assumption about
the content, and it is left free to be filled with custom metadata. The only
requirement is to set this filed with zeroes when is not used.

@@

\section{5G}
The continuous innovations in the mobile network connectivity is leading to the
creation of a new standard, the 5G, which is estimated to arrive to the market
consumer in 2020~\cite{iwamura2015ngmn}. Lead by the Next Generation Mobile
Network (NGMN) alliance, a group composed by the major players in the field of
mobile connectivity, the 5G aims to offer not only at the end-user a new way to
browse the web, download and watch interactive content but also to create an
ad-hoc solution for Machine-to-Machine (M2M) data traffic, which is increasing
more than ever thanks to the spreading of IoT devices, which sensors need to
continuously send data to servers/data-centers. Relatively to \emph{Long Term
  Evolution} (LTE), 5G points to have data rates $10$ times better, with $10$
times smaller end-to-end latency and an increased connection density by $100$
times~\cite{alliance20155g}.
